Induced universal graphs for families of small graphs

### 2021-10-25 paper/paper.tex
Tidy
More about trees

This paper uses a brute-force approach similar to that of Preen to find minimal
induced universal graphs for families of small graphs.  Our contributions
are (1) an ordering strategy to reduce the number of subgraph isomorphism
calls required by the brute force algorithm; (2) exact values for $f(k)$ $(5 \leq k \leq 6)$
and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
search algorithm to find small (but not necessarily optimal) induced universal graphs
(5) exact results for families of small trees.

The sequel is structured as follows.
Section~\ref{sec:method} describes the basic brute force method.
Section~\ref{sec:iteration-order} compares four methods for sorting
the graphs in $\calF$, and compares their effect on the number of subgraph
isomorphism calls made by the brute-force algorithm.
Sections~\ref{sec:results5} presents exact
values of $f(k)$ and $F(k)$ for $k \leq 5$.
Section~\ref{sec:f6} gives the value of $f(6)$ and and
Section~\ref{sec:f7} gives bounds on $f(7)$; in each of these sections, the lower
bound is proven and a graph obtained by heuristic search demonstrating the
upper bound is given.  Finally, Section~\ref{sec:trees} gives results for
families of trees.

@@

Given $\calF$ and $n$,
we use the brute-force approach
shown in \Cref{alg:brute-force} --- which is essentially the same as the method
described by Preen \cite{preen_math_se} ---
to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$.
The function $\FuncSty{AllInducedUniversalGraphs}()$ is the entry point.
\Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
of $n$-vertex graphs, and adds to the collection $\calG$ those that are induced universal
for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
whether a graph $G$ is induced universal for $\calF$ by checking for
each $H \in \calF$ that $H$ is isomorphic to a subgraph of $G$.

@@

\begin{figure}[htb]
\centering
\footnotesize
\verb|0 1 1 1 1 1 0 1 1 0 0 0 1 0| \\
\verb|1 0 1 1 1 1 1 1 1 1 1 0 0 1| \\
\verb|1 1 0 1 1 1 0 1 1 0 1 0 1 0| \\
\verb|1 1 1 0 1 1 1 0 0 0 1 1 0 0| \\
\verb|1 1 1 1 0 1 0 1 0 0 0 0 0 0| \\
\verb|1 1 1 1 1 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 0 1 0 0 0 0 0 0 0 1 1 1| \\
\verb|1 1 1 0 1 0 0 0 0 0 0 1 1 0| \\
\verb|1 1 1 0 0 0 0 0 0 0 0 0 1 0| \\
\verb|0 1 0 0 0 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 1 1 0 0 0 0 0 0 0 1 1 0| \\
\verb|0 0 0 1 0 1 1 1 0 1 1 0 0 1| \\
\verb|1 0 1 0 0 0 1 1 1 0 1 0 0 0| \\
\verb|0 1 0 0 0 1 1 0 0 1 0 1 0 0|
\caption{The adjacency matrix of a 14-vertex induced universal graph for the family of all
six-vertex graphs}
\label{fig:adjmat14}
\end{figure}

@@

\begin{figure}[htb]
\centering
\footnotesize
\verb|0 1 1 1 1 1 1 1 1 1 0 1 1 1 1 1 0 0| \\
\verb|1 0 1 1 1 1 1 0 1 1 1 0 1 1 0 1 0 0| \\
\verb|1 1 0 1 1 1 1 1 0 1 0 1 0 0 0 0 0 1| \\
\verb|1 1 1 0 1 1 1 0 1 0 0 1 0 0 0 0 0 1| \\
\verb|1 1 1 1 0 1 1 1 1 0 0 0 0 0 1 0 0 0| \\
\verb|1 1 1 1 1 0 1 0 1 0 0 0 0 1 1 1 1 1| \\
\verb|1 1 1 1 1 1 0 0 0 0 0 0 0 1 0 0 0 1| \\
\verb|1 0 1 0 1 0 0 0 0 0 0 0 0 0 1 1 1 0| \\
\verb|1 1 0 1 1 1 0 0 0 0 0 0 0 1 1 1 0 1| \\
\verb|1 1 1 0 0 0 0 0 0 0 0 0 0 0 1 1 1 1| \\
\verb|0 1 0 0 0 0 0 0 0 0 0 0 0 1 0 0 1 0| \\
\verb|1 0 1 1 0 0 0 0 0 0 0 0 0 0 1 1 1 0| \\
\verb|1 1 0 0 0 0 0 0 0 0 0 0 0 0 0 1 1 1| \\
\verb|1 1 0 0 0 1 1 0 1 0 1 0 0 0 0 0 0 1| \\
\verb|1 0 0 0 1 1 0 1 1 1 0 1 0 0 0 0 0 1| \\
\verb|1 1 0 0 0 1 0 1 1 1 0 1 1 0 0 0 1 1| \\
\verb|0 0 0 0 0 1 0 1 0 1 1 1 1 0 0 1 0 1| \\
\verb|0 0 1 1 0 1 1 0 1 1 0 0 1 1 1 1 1 0|
\caption{The adjacency matrix of an 18-vertex induced universal graph for the family of all
seven-vertex graphs}
\label{fig:adjmat18}
\end{figure}

@@

In order to generate induced universal graphs for families of
all trees on $k$ vertices, we ran \Cref{alg:brute-force}
using lists of trees published by Brendan
McKay.\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/trees.html}}

@@

@@

\begin{table}[h!]
\centering
\begin{tabular}{r r r}
 \toprule
 $k$ & $t(k)$ & $T(k)$ \\ [0.5ex]
 \midrule
 1 & 1 & 1 \\
 2 & 2 & 1 \\
 3 & 3 & 1 \\
 4 & 5 & 2 \\
 5 & 7 & 18 \\
 6 & 9 & 66 \\
 \bottomrule
\end{tabular}
\caption{For each $k$, $t(k)$ is the minimum order of a graph containing all $k$-vertex trees as
induced subgraphs, and $T(k)$ is the number of distinct $t(k)$-vertex graphs that contain
all $k$-vertex trees as induced subgraphs.}
\label{tab:treeresults}
\end{table}

>>>>

This paper uses a brute-force approach similar to that of Preen to find minimal
induced universal graphs for families of small graphs.  Our contributions
are (1) an ordering strategy to reduce the number of subgraph isomorphism
calls required by the brute force algorithm; (2) exact values for $f(k)$ $(5 \leq k \leq 6)$
and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
search algorithm to find small (but not necessarily optimal) induced universal graphs
(5) exact values of the order of a minimal induced universal graph
for the families of all trees on $k$ vertices, for $1 \leq k \leq 8$.

The sequel is structured as follows.
Section~\ref{sec:method} describes the basic brute force method.
Section~\ref{sec:iteration-order} compares four methods for sorting
the graphs in $\calF$, and compares their effect on the number of subgraph
isomorphism calls made by the brute-force algorithm.
Sections~\ref{sec:results5} presents exact
values of $f(k)$ and $F(k)$ for $k \leq 5$.
Section~\ref{sec:f6} gives the value of $f(6)$ and and
Section~\ref{sec:f7} gives bounds on $f(7)$; in each of these sections, the lower
bound is proven and a graph obtained by heuristic search demonstrating the
upper bound is given.  Finally, Section~\ref{sec:trees} describes a specialised
algorithm for families of trees, and gives results for
these families.

@@

Given $\calF$ and $n$,
we use the brute-force approach
shown in \Cref{alg:brute-force} --- which is essentially the same as the method
described by Preen \cite{preen_math_se} ---
to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$.
The entry point is the function $\FuncSty{AllInducedUniversalGraphs}()$.
\Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
of $n$-vertex graphs, and adds to the collection $\calG$ those that are induced universal
for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
whether a graph $G$ is induced universal for $\calF$ by checking for
each $H \in \calF$ that $H$ is isomorphic to a subgraph of $G$.

@@

\begin{figure}[htb]
\centering
\scriptsize
\verb|0 1 1 1 1 1 0 1 1 0 0 0 1 0| \\
\verb|1 0 1 1 1 1 1 1 1 1 1 0 0 1| \\
\verb|1 1 0 1 1 1 0 1 1 0 1 0 1 0| \\
\verb|1 1 1 0 1 1 1 0 0 0 1 1 0 0| \\
\verb|1 1 1 1 0 1 0 1 0 0 0 0 0 0| \\
\verb|1 1 1 1 1 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 0 1 0 0 0 0 0 0 0 1 1 1| \\
\verb|1 1 1 0 1 0 0 0 0 0 0 1 1 0| \\
\verb|1 1 1 0 0 0 0 0 0 0 0 0 1 0| \\
\verb|0 1 0 0 0 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 1 1 0 0 0 0 0 0 0 1 1 0| \\
\verb|0 0 0 1 0 1 1 1 0 1 1 0 0 1| \\
\verb|1 0 1 0 0 0 1 1 1 0 1 0 0 0| \\
\verb|0 1 0 0 0 1 1 0 0 1 0 1 0 0|
\caption{The adjacency matrix of a 14-vertex induced universal graph for the family of all
six-vertex graphs}
\label{fig:adjmat14}
\end{figure}

@@

\begin{figure}[htb]
\centering
\scriptsize
\verb|0 1 1 1 1 1 1 1 1 1 0 1 1 1 1 1 0 0| \\
\verb|1 0 1 1 1 1 1 0 1 1 1 0 1 1 0 1 0 0| \\
\verb|1 1 0 1 1 1 1 1 0 1 0 1 0 0 0 0 0 1| \\
\verb|1 1 1 0 1 1 1 0 1 0 0 1 0 0 0 0 0 1| \\
\verb|1 1 1 1 0 1 1 1 1 0 0 0 0 0 1 0 0 0| \\
\verb|1 1 1 1 1 0 1 0 1 0 0 0 0 1 1 1 1 1| \\
\verb|1 1 1 1 1 1 0 0 0 0 0 0 0 1 0 0 0 1| \\
\verb|1 0 1 0 1 0 0 0 0 0 0 0 0 0 1 1 1 0| \\
\verb|1 1 0 1 1 1 0 0 0 0 0 0 0 1 1 1 0 1| \\
\verb|1 1 1 0 0 0 0 0 0 0 0 0 0 0 1 1 1 1| \\
\verb|0 1 0 0 0 0 0 0 0 0 0 0 0 1 0 0 1 0| \\
\verb|1 0 1 1 0 0 0 0 0 0 0 0 0 0 1 1 1 0| \\
\verb|1 1 0 0 0 0 0 0 0 0 0 0 0 0 0 1 1 1| \\
\verb|1 1 0 0 0 1 1 0 1 0 1 0 0 0 0 0 0 1| \\
\verb|1 0 0 0 1 1 0 1 1 1 0 1 0 0 0 0 0 1| \\
\verb|1 1 0 0 0 1 0 1 1 1 0 1 1 0 0 0 1 1| \\
\verb|0 0 0 0 0 1 0 1 0 1 1 1 1 0 0 1 0 1| \\
\verb|0 0 1 1 0 1 1 0 1 1 0 0 1 1 1 1 1 0|
\caption{The adjacency matrix of an 18-vertex induced universal graph for the family of all
seven-vertex graphs}
\label{fig:adjmat18}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \footnotesize
%    \tt
       \begin{tikzpicture}[>=latex',line join=bevel,scale=.6]
        \tikzstyle{every node}=[]
          \input{img/regions-for-trees}
      \end{tikzpicture}
    \caption{The adjacency-matrix regions used in the 
        ``symmetry breaking completion'' method. For this example, $n=11$ and $k=7$.}
\label{fig:regions-for-trees}
\end{figure}

To generate induced universal graphs for families of
all trees on $k$ vertices, we tried two approaches.
The first approach was to run \Cref{alg:brute-force}, with the family
$\calF$ of order-$k$ trees loaded from
a list of trees published by Brendan
McKay.\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/trees.html}}
The second approach, described in the following subsection, generates
only candidate graphs that contain a star --- one of the elements
of $\calF$ ---as an induced subgraph.  The final part of this section
gives results generated using these two methods, along with additional
upper bounds generated using a specialised version of our hill-climbing
algorithm.

\subsection{The completion method}

Our second approach was to embed the star $S_{k-1}$ --- which must be
present since it is a tree on $k$ vertices --- in the top left position
of the adjacency matrix, then systematically try all possible ways to complete
the adjacency matrix.  For each completed adjacency matrix, we test whether
the graph contains as an induced subgraph each tree in our family.
If it does, and if the graph is not isomorphic to any induced universal
graph that has already been found, we add it to our collection of induced
universal graphs.
We will refer to this as the ``naive completion'' method.

Because the naive completion method has to test a huge number of graphs,
many pairs of which are isomorphic, we created an improved version of the
algorithm, which we will refer to as the ``symmetry breaking completion''
method.  This adds two symmetry-breaking constraints on the adjacency matrix.
The first of these constraints concerns the grey shaded region at the bottom-right
of the adjacency matrix, which has $n - k$ rows and columns.  Rather than
trying all $2^{n-k \choose 2}$ possible entries for this region of the
adjacency matrix, we use a list containing a canonical adjacency matrix
for each possible graph of order $n-k$, and we require that the shaded
region be equal to one of these adjacency matrices.

The second symmetry-breaking constraint concerns the yellow shaded region
in \Cref{fig:regions-for-trees}. Each row of this region has $n-k$ entries,
named $x_{i1}, \dots, x_{i,n-k}$.  For our symmetry-breaking constraint, we
view each row as the binary number $x_{i1}, \dots, x_{i,n-k}$
(with $x_{i,n-k}$ as the least significant digit), and we insist
that these numbers are in nondecreasing order; thus
$x_{21}, \dots, x_{2,n-k} \leq \dots \leq x_{k1}, \dots, x_{k,n-k}$.
In other words, we impose a lexicographic ordering constraint
on the rows of this region.

The algorithm with these symmetry-breaking constraints is shown
in \Cref{alg:completion}.  The call to $\FuncSty{MakeGraph}()$ on
\lineref{MakeGraphLine} constructs a graph by building its
adjacency matrix as shown in \Cref{fig:regions-for-trees}.  The
first $k$ vertices are connected in the form of a star;
for each $i$ such that $1 \leq i \leq j$, the digits of the binary
representation of $x_i$
are assigned to $x_{i1}, \dots, x_{i,n-k}$, and these values
are placed in the adjacency matrix as shown in the figure.  Finally,
the small adjacency matrix $M$ is copied to the bottom-right of
the adjacency matrix of the constructed graph.

For the collection of graphs $\calG$ defined on \lineref{DefineSetG},
isomorphic graphs are considered identical.  Thus, we only add a graph
to $\calG$ if it is not isomorphic to any existing member of the
collection.\footnote{An efficient method to ensure that isomorphic
graphs are not stored in $\calG$ would be label the vertices
of a graph $G$ canonically before adding it to the collection
\cite{DBLP:journals/jsc/McKayP14}, and to store the collection as a
hash set.  Because this part of the prograph is not a performance
bottleneck, and to miminise software dependencies, our implementation
uses the simpler approach of testing whether $G$ is isomorphic
to each existing member of $\calG$, in turn.}

\begin{algorithm}[h!]
\DontPrintSemicolon
\nl $\FuncSty{Search}(n,k,(x_1, \dots, x_j))$ \;
\nl \KwData{Natural numbers $n$ and $k$, and a sequence
    $(x_1, \dots, x_j))$ of nonnegative integers such that the binary
    representation of each $x_i$ represents entries $x_{i1}, \dots, x_{i,n-k}$
    of an adjacency matrix, as shown in \Cref{fig:regions-for-trees}}
\nl \KwResult{The set of all order-$n$ graphs that are induced universal
    for the family of trees of order $k$, and have an adjacency matrix
    given by $\FuncSty{MakeGraph}(n,k,X,M)$ for some completion
    $X$ of $(x_1, \dots, x_j)$ and some $M \in \mathcal{M}(n-k)$}
\nl \Begin{
\nl    $\calG \gets \emptyset$ \label{DefineSetG} \;
\nl    \If {$j \leq 1$}{
\nl      \For {$i \in \{0, \dots, 2^{n-k} - 1\}$}{
\nl        $\calG \gets \calG \cup \FuncSty{Search}(n,p,(x_1, \dots, x_j, i))$
  }
}
\nl    \ElseIf {$j \leq k$}{
\nl      \For {$i \in \{0, \dots, x_j\}$}{
\nl        $\calG \gets \calG \cup \FuncSty{Search}(n,p,(x_1, \dots, x_j, i))$
  }
}
\nl    \Else {
\nl      \For {$M \in \mathcal{M}(n-k)$}{
\nl        $G \gets \FuncSty{MakeGraph}(n, k, (x_1, \dots, x_j), M)$ \label{MakeGraphLine}\;
\nl        \If {$G$ contains every tree of order $k$ as an induced subgraph}{
\nl          $\calG \gets \calG \cup \{G\}$ \;
  }
}
}
\nl    $\KwSty{return}$~$\calG$ \;
    }
\medskip
\nl $\FuncSty{Start}(n,k)$ \;
\nl \KwData{Natural numbers $n$ and $k$ such that $k<n$}
\nl \KwResult{The set of all graphs of order $n$ that are induced universal for
    the family of trees of order $k$.}
\nl \Begin{
\nl    $\KwSty{return}$~$\FuncSty{Search}(n,k,())$ \;
    }
\caption{The symmetry-breaking completion algorithm for finding all graphs of order $n$
    that are induced universal for the family of all order-$k$ trees}
\label{alg:completion}
\end{algorithm}

Because of the following proposition, the symmetry breaking completion
method finds, up to isomorphism, all induced universal graphs for the family
of trees on $k$ vertices.

\begin{proposition}
    Let $n, k$ be given, and let $G$ be a graph on $n$ vertices that is
    induced universal for the family of trees of order $k$.  Let $\mathcal{M}$
    be a set of canonical adjacency matrices for $\calF(n-k)$.  There is
    a graph on $n$ vertices numbered $1, \dots, n$ that is isomorphic
    to $G$ and satisfies the two symmetry-breaking constraints described
    above.
\end{proposition}
\begin{proof}
    Since $G$ is induced universal for the family of trees of order $k$, it
    contains the star $S_{k-1}$ as an induced subgraph.  Arbitrarily
    choose one such star in $G$ and give the label 1 to its vertex of degree
    $k-1$.  Its remaining vertices will be given the labels $\{2, \dots, k\}$,
    in an order we will specify later.
    The remaining $n-k$ vertices of $G$ induce a subgraph isomorphic
    to one of the adjacency matrices in $\mathcal{M}$; call this matrix $M$.
    By choosing an appropriate relabelling of the $n-k$ vertices of $G$,
    the region of the adjacency matrix corresponding to these vertices
    can be made equal to $M$.  Thus, the first symmetry-breaking constraint
    is satisfied.

    The second symmetry-breaking constraint can now be satisfied by assigning
    labels $\{2, \dots, k\}$ to the degree-1 vertices of the star $S_{k-1}$
    in such an order that their adjacencies to the vertices outside the star,
    viewed as binary numbers, are in lexicographic order.
\end{proof}

\subsection{Results for families of trees}

The symmetry-breaking completion method is much faster than \Cref{alg:brute-force}.
For $n=9$ and $k=6$, for example, the completion method takes 9 seconds whereas
the brute-force algorithm takes 46 seconds. The difference in run-times becomes
even greater for larger values of $n$ and $k$.
Using the symmetry-breaking completion method,
we were able to find the values of $t(k)$ and $T(k)$ for $k \leq 7$. We were also
able to show that $t(8) > 12$.  We confirmed
the results for $k \leq 6$ using \Cref{alg:brute-force}.

@@

To obtain upper bounds on $t(k)$ for larger values of $k$, we use a version of
\Cref{alg:heuristic} with minor modifications.  Instead of initially embedding
a clique and an independent set as in \Cref{fig:heuristic-regions}, we embed
a single tree --- the $k$-vertex star $S_{k-1}$ --- as shown in
\Cref{fig:heuristic-regions-for-trees}.  Each possible edge marked with a dot
in the figure is added with probability 0.1.

\begin{figure}[h!]
    \centering
    \footnotesize
%    \tt
       \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \tikzstyle{every node}=[]
          \input{img/heuristic-regions-for-trees}
      \end{tikzpicture}
    \caption{The adjacency matrix used in our heuristic algorithm when searching
        for an induced universal graph of order 9 for the set of all trees
        of order 6.  The blue
        region, a star with 6 vertices, remains unchanged as the algorithm runs.
        The remaining possible edges, marked $\boldsymbol{\cdot}$,
        are flipped between 0 and 1 as the algorithm progresses.}
\label{fig:heuristic-regions-for-trees}
\end{figure}

Using this heuristic algorithm, we were able to find induced universal graphs
to demonstrate that $t(8) \leq 13$, $t(9) \leq 15$ and $t(10) \leq 18$.
Since we have matching lower and upper bounds for $t(8)$, we can state
that $t(8) = 13$.

@@

\begin{table}[h!]
\centering
\begin{tabular}{r r r}
 \toprule
 $k$ & $t(k)$ & $T(k)$ \\ [0.5ex]
 \midrule
 1 & 1 & 1 \\
 2 & 2 & 1 \\
 3 & 3 & 1 \\
 4 & 5 & 2 \\
 5 & 7 & 18 \\
 6 & 9 & 66 \\
 7 & 11 & 687 \\
 \bottomrule
\end{tabular}
\caption{For each $k$, $t(k)$ is the minimum order of a graph containing all $k$-vertex trees as
induced subgraphs, and $T(k)$ is the number of distinct $t(k)$-vertex graphs that contain
all $k$-vertex trees as induced subgraphs.}
\label{tab:treeresults}
\end{table}
